Gradual Soundness: Lessons from Static Python

### 2022-05-18 pj-7.1/paper.tex
pj: bookkeeping
- add access dates to ALL urls
- doi or url for all bib entries

\overfullrule=1mm

@@

\author[a]{Kuang-Chen Lu}
\authorinfo{(\email{LuKuangchen1024@gmail.com}) is a PhD student at Brown University.}
\affiliation[a]{Brown University, Providence, RI, USA}
\author[a]{Ben Greenman}
\authorinfo{(\email{benjamin.l.greenman@gmail.com}) is a postdoc at Brown University.}
\author[b]{Carl Meyer}
\authorinfo{(\email{carljm@fb.com}) is a software engineer on the Python Language \& Runtime team at Meta.}
\affiliation[b]{Meta, Menlo Park, CA, USA}
\author[b]{Dino Viehland}
\authorinfo{(\email{dinoviehland@fb.com}) is a software engineer at Meta working on the Cinder Python runtime. Previously he has worked on other Python related efforts including Python Tools for Visual Studio and IronPython.}
\author[b]{Aniket Panse}
\authorinfo{(\email{aniketpanse@fb.com}) is a software engineer on the Python Language \& Runtime team at Meta.}
\author[a]{Shriram Krishnamurthi}
\authorinfo{(\email{shriram@gmail.com}) is the Vice President of Programming Languages (no, not really) at Brown University.}

@@

\SP{} represents a new approach to the design of gradually-typed
languages.
Whereas prior works choose to either force changes upon untyped code
with \emph{concrete} types~\cite{wnlov-popl-2010,mt-oopsla-2017}
or accommodate untyped code as-is with \emph{migratory}
types~\cite{st-sfp-2006,tfffgksst-snapl-2017}, \SP{} provides a multi-level
type system that supports both styles without expensive run-time checks.
Shallow types accommodate Python code, but offer little in the way
of type-based reasoning and type-directed optimization.
Concrete types and primitive types give stronger type guarantees, but have
drawbacks:
concrete types force changes upon untyped code and primitive types
add restrictions to typed code.
The overall system, in which programmers can choose between untyped code
and several increasingly-restrictive typed languages,
is a combination of gradual typing and progressive
types~\cite{pqk-onward-2012} that we call \emph{gradual soundness}.

@@

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
assertions, loops, exception handlers, and delete statements.
It follows \SP{}'s approach to typing these features.
The model is available at:
\begin{center}
  \url{https://github.com/brownplt/insta-model}
\end{center}
Because there is no prior formalization of (all of) Python and of \SP{}, we cannot
\emph{verify} that the model matches them. Instead, we have applied
three methods to give confidence that our model matches reality.
First, we manually reviewed the model in depth---using our
own expertise---to look for non-conformance.  Second, 
we used Redex's property-based testing tools~\cite{kf-sfp-2009} to check
type soundness.
Specifically, we used Redex to generate well-formed terms (using its default
algorithm) and checked whether the well-typed terms that terminated within a
fixed time budget (1,200 expressions and 11,000 programs in a typical run)
produced only well-typed values and allowed errors.
% \footnote{Property-based testing helped us find and fix several issues in the
% model.  None of issues had implications for \SP{}'s soundness.}
Finally, we employed the well-established method of testing end-to-end
conformance with a test suite~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}.
We translated \numSPtest{} tests from the \SP{} regression suite to the
syntax of the model and confirmed that the results
do match,
which suggests that the model conforms to actual \SP{}.

@@

Concrete types originated in Thorn~\cite{bfnorsvw-oopsla-2009,wnlov-popl-2010}
and StrongScript~\cite{rzv-ecoop-2015}, two early languages that support typed
and untyped objects.
Thorn is a scripting language for the JVM; StrongScript is designed to
interoperate with JavaScript.
Concrete types came to \SP{} by way of Nom~\cite{mt-oopsla-2017} and its
successor MonNom~\cite{mt-oopsla-2021}.
The impressive performance figures in the original Nom paper inspired the \SP{}
team to adapt the method.
But whereas MonNom has been exploring designs for generic concrete types that
support (a flexible form of) the gradual guarantees~\cite{svcb-snapl-2015},
\SP{} opts for a simpler and more rigid design~(\cref{s:type-dynamic}).

@@

The 26 GitHub issues that arose from our work may be found at the following three links.
A list of issue numbers follows each link.

@@

Benchmarks \bmname{richards} and \bmname{fannkuch} ran on the following commit:

\begin{center}
\footnotesize\url{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}
\end{center}

\noindent\bmname{deltablue} ran on a later commit:

\begin{center}
\footnotesize\url{https://github.com/facebookincubator/cinder/commit/c9d14c4474facfeca15b618015b58b99a3e86d25}
\end{center}

\noindent\bmname{nbody} ran on an even later commit:

\begin{center}
\footnotesize\url{https://github.com/facebookincubator/cinder/commit/144a21e66aeae5cde08847482249396b2e16eccc}
\end{center}

>>>>

%\overfullrule=1mm

@@

\author[a]{Kuang-Chen Lu}
\authorinfo{(\email{LuKuangchen1024@gmail.com}) is a PhD student at Brown University.}
\affiliation[a]{Brown University, Providence, RI, USA}
\author[a]{Ben Greenman}
\authorinfo{(\email{benjamin.l.greenman@gmail.com}) is a postdoc at Brown University.}
\author[b]{Carl Meyer}
\authorinfo{(\email{carljm@fb.com}) is a software engineer on the Python Language \& Runtime team at Meta.}
\affiliation[b]{Meta, Menlo Park, CA, USA}
\author[b]{Dino Viehland}
\authorinfo{(\email{dinoviehland@fb.com}) is a software engineer at Meta working on the Cinder Python runtime. Previously he has worked on other Python related efforts including Python Tools for Visual Studio and IronPython.}
\author[b]{Aniket Panse}
\authorinfo{(\email{aniketpanse@fb.com}) is a software engineer on the Python Language \& Runtime team at Meta.}
\author[a]{Shriram Krishnamurthi}
\authorinfo{(\email{shriram@brown.edu}) is the Vice President of Programming Languages (no, not really) at Brown University.}

@@

\SP{} represents a new approach to the design of gradually-typed
languages.
Whereas prior works choose to either force changes upon untyped code
with \emph{concrete} types~\cite{wzlov-popl-2010,mt-oopsla-2017}
or accommodate untyped code as-is with \emph{migratory}
types~\cite{st-sfp-2006,tfffgksst-snapl-2017}, \SP{} provides a multi-level
type system that supports both styles without expensive run-time checks.
Shallow types accommodate Python code, but offer little in the way
of type-based reasoning and type-directed optimization.
Concrete types and primitive types give stronger type guarantees, but have
drawbacks:
concrete types force changes upon untyped code and primitive types
add restrictions to typed code.
The overall system, in which programmers can choose between untyped code
and several increasingly-restrictive typed languages,
is a combination of gradual typing and progressive
types~\cite{pqk-onward-2012} that we call \emph{gradual soundness}.

@@

To validate the design of \SP{}, we developed a model of the language in PLT
Redex~\cite{kcdeffmrtf-popl-2012}.
The model covers a substantial part of the Python language, including
assertions, loops, exception handlers, and delete statements.
It follows \SP{}'s approach to typing these features.
The model is on GitHub (accessed 2022-05-17):
\begin{center}
  \url{https://github.com/brownplt/insta-model}
\end{center}
Because there is no prior formalization of (all of) Python and of \SP{}, we cannot
\emph{verify} that the model matches them. Instead, we have applied
three methods to give confidence that our model matches reality.
First, we manually reviewed the model in depth---using our
own expertise---to look for non-conformance.  Second, 
we used Redex's property-based testing tools~\cite{kf-sfp-2009} to check
type soundness.
Specifically, we used Redex to generate well-formed terms (using its default
algorithm) and checked whether the well-typed terms that terminated within a
fixed time budget (1,200 expressions and 11,000 programs in a typical run)
produced only well-typed values and allowed errors.
% \footnote{Property-based testing helped us find and fix several issues in the
% model.  None of issues had implications for \SP{}'s soundness.}
Finally, we employed the well-established method of testing end-to-end
conformance with a test suite~\cite{gsk-ecoop-2010,gclpk-dls-2012,pmmwplck-oopsla-2013,bcfgmnss-popl-2014,fgpssmds-popl-2016}.
We translated \numSPtest{} tests from the \SP{} regression suite to the
syntax of the model and confirmed that the results
do match,
which suggests that the model conforms to actual \SP{}.

@@

Concrete types originated in Thorn~\cite{bfnorsvw-oopsla-2009,wzlov-popl-2010}
and StrongScript~\cite{rzv-ecoop-2015}, two early languages that support typed
and untyped objects.
Thorn is a scripting language for the JVM; StrongScript is designed to
interoperate with JavaScript.
Concrete types came to \SP{} by way of Nom~\cite{mt-oopsla-2017} and its
successor MonNom~\cite{mt-oopsla-2021}.
The impressive performance figures in the original Nom paper inspired the \SP{}
team to adapt the method.
But whereas MonNom has been exploring designs for generic concrete types that
support (a flexible form of) the gradual guarantees~\cite{svcb-snapl-2015},
\SP{} opts for a simpler and more rigid design~(\cref{s:type-dynamic}).

@@

The 26 GitHub issues that arose from our work may be found at the following three links.
A list of issue numbers follows each link (all accessed 2022-05-17).

@@

\begin{itemize}
  \item
    Benchmarks \bmname{richards} and \bmname{fannkuch} ran on commit
    \href{https://github.com/facebookincubator/cinder/commit/d0d071a9acf3e65700e7c6f8982c5087c700d116}{d0d071}
  \item
    \bmname{deltablue} ran on commit
    \href{https://github.com/facebookincubator/cinder/commit/c9d14c4474facfeca15b618015b58b99a3e86d25}{c9d14c}
  \item
    \bmname{nbody} ran on commit
    \href{https://github.com/facebookincubator/cinder/commit/144a21e66aeae5cde08847482249396b2e16eccc}{144a21}
\end{itemize}
